Large-scale spatiotemporal variation in vital rates and population dynamics of an alpine bird

### 2024-09-13 Manuscript_Nater_etal.qmd
Added FraFra to author list
Updates to manuscript text

---
title: "Large-scale spatiotemporal variation in vital rates and population dynamics of an alpine bird"
format: html
author: Chloé R. Nater, James A. Martin, Erlend B. Nilsen
editor: visual
bibliography: rypeIDSM_MS2.bib
---

@@

Reproducibility and ease of repeating analyses was a key focus when developing the multi-area IDSM. To that end, we built a function-based workflow (@fig-workflow) that includes a variety of options for controlling modelling decisions such as the year range of data to consider, the level of spatial aggregation (i.e. reporting district vs. survey locality), whether to model time variation in survival and/or effects of rodent occupancy, whether to run MCMC chains sequentially or in parallel, etc. We then set up three pipelines for executing the workflow that differ in their degree of automation and user interface to meet different needs and resource constraints. The first is a step-by-step R script to be run manually that is suitable for exploration, development, and debugging. The second is a largely automated "R targets" pipeline [@landau2021], which allows executions through a single command and maximizes efficiency by keeping track of the "up-to-date" vs. "outdated" status of different steps in the workflow. The third is a pipeline that is run directly from command line, sets up a fully reproducible environment through Nix [nix2004], and parallelizes the MCMC outside of R using GNUparallel [@tange2024]. This option avoids a range of issues that can arise with R's internal parallelization (e.g. processes running even when the parent R session has been restarted, hard to debug, bad error handling, etc.) and is particularly well suited for running on servers and high-performance computing clusters. For more details on pipeline implementations and options, we refer the reader to the vignettes in the GitHub repository: <https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp>.

@@

Average population growth rates over the last four years (2018-2021) ranged from moderate declines (`r lambda.min$post.sum` in the "`r lambda.min$area.name`" area) to \> 50% increase (`r lambda.max$post.sum` in the "`r lambda.max$area.name`" area). In the majority of reporting areas (`r n.pos.lambda` out of 41), populations of willow ptarmigan have been increasing over the period 2018-2021 (@fig-lam). Some areas -- predominantly in central Norway -- also had declining populations, but many of those declines followed upon periods of increase between the start of data collection in 2007 and sometime between 2016 and 2018 (SF "TimeSeries_popDens1.pdf").

@@

The highest recent population growth rates were estimated for areas with relatively low recent population densities across latitudes but we did not find evidence for a strong association between population growth rates and population densities overall ([@fig-paramCorrs] A).

@@

## Survival probabilities and recruitment rates

@@

The MCMC chains for many of the area-specific average survival probabilities and recruitment rates, as well as for the global averages for both vital rates, were mixing rather poorly. Despite that, mixing was good and resulting posteriors well defined for the area- and year-specific estimates of survival and recruitment (SF "PostDens_tS_tR.pdf"). There was substantial variation in both vital rates across time (SFs "TimeSeries_pSurv.pdf" and "TimeSeries_rRep.pdf"). In a substantial number of areas, the years 2011, 2014, and 2018 not only supported high population densities (see above) but were also characterized by both high recruitment and low subsequent survival. <!-- ERLEND: Here and above: I would perhaps consider deleting the reference to 2011 as a "high density year", as there is really a lot of spatial variation in this pattern. I'm not so convinced that is the case in general. CHLOE: I checked again in my "scoring sheet", and it's a clear density peak in 16 areas, which is not far from 2014 (19 areas) and 2018 (18 areas). Obviously there is variation, but I do think this is substantial. I would like to leave it in, but have changed the phrasing to "a substantial number of areas" instead of "many"/"a large number". Okay? ERLEND: Ok--> The overall low density years 2012 and 2015, conversely, often featured lower recruitment and, in some cases, higher survival. Notably, there were also years with very little spatial synchrony, i.e. very different relative yearly survival probabilities and recruitment rates (e.g. 2010 and 2020 for survival and 2013, 2016, and 2017 for recruitment).

@@

The model predicted a positive global effect of rodent occupancy on recruitment rate (average slope on the log scale = `r betaR.hyper.mean$post.sum`). Nonetheless, spatial variation in the rodent effect was substantial (random effect SD = `r betaR.sd.spatial$post.sum`). This resulted in negative (median) effects in `r n.neg.betaR` areas, positive (median) effects in `r n.pos.betaR` areas, and a range of effect sizes from `r betaR.min$post.sum` (area "`r betaR.min$area.name`") to `r betaR.max$post.sum` (area "`r betaR.max$area.name`", @fig-paramCorrs C, SF "Rep_betaR.R.pdf"). The largest positive rodent effects were estimated for areas in the very North of Norway, as well as in the mountaineous regions in the central and southwestern parts of the country (SF "betaR_Map.pdf"). Effects with negative posterior medians were located mostly at intermediate latitudes, but we note that all of these had posterior distributions featuring substantial overlap with 0 (@fig-paramCorrs C).

@@

Detection decay parameters, which determine detection probability in distance sampling surveys, varied across areas from between `r Det.min$post.sum` in "`r Det.min$area.name`" to `r Det.max$post.sum` in "`r Det.max$area.name`, resulting in detection probabilities over the transect sites ranging from `r pDet.min$post.sum` to `r pDet.max$post.sum`, respectively (truncation distance = 200 m) . The global average detection decay was `r Det.hyper.mean$post.sum` (detection probability = `r pDet.hyper.mean$post.sum`), and in general, higher values were more common in the Southern half of the country than the Northern half (SF"Avg_detect_Map.pdf"). Variation in detection over time was modest on average but the degree of temporal changes varied by area, with some areas having nearly constant detection while others showed variation by factors larger than 1.5 (SF "TimeSeries_pDetect.pdf"). The estimated average among-year variation in detection decay (random effect log SD = `r Det.sd.temporal$post.sum`) was almost identical in magnitude to spatial (`r Det.sd.spatial$post.sum`) and residual (`r Det.sd.residual$post.sum`) variation.

@@

The relative importance of different components for explaining parameter variation differed among recruitment rate, survival probability, and detection decay (@fig-varDecomp). The largest portion of variation in recruitment was attributed to residual variation (`r Rec_R` %), followed by spatial (`r Rec_A` %) and temporal (`r Rec_Y` %) variation. Rodent occupancy, which contains both a spatial and a temporal dimension, explained `r Rec_Cov` % of the total variation. For survival, there was large uncertainty in the estimated proportions of variance explained by different components. The model predicted similar potential contributions from temporal (`r Surv_Y` %) and residual (`r Surv_R` %) variation and suggested that spatial variation was only responsible for `r Surv_A` % of the total variance. Total variance in detection decay was attributed evenly to spatial, temporal, and residual variation at `r Det_A` %, `r Det_Y` %, and `r Det_R` %, respectively.

@@

Ptarmigan densities across the 41 reporting districts included in our analyses varied from around 2 birds/km^2^ to 55 birds/km^2^, with the lowest densities occurring far north in the country, as well as on the west coast and in the mountains in central Norway [@fig-densA]. The same spatial pattern was also evident at the level of the demographic rates: consistent with basic life history theory [@stearns1992], average recruitment rates were inversely related to average survival probabilities [@fig-paramCorrs B], and the slower life histories (higher survival and lower recruitment) were more common in the northern and mountaineous parts of the country. This aligns with previous studies reporting relatively slower bird life histories in alpine / high altitude areas [e.g. @sandercock2005; @bears2009; @Wilson_Martin_2011; @alice2016]. In Norway, the northern and mountainous areas are characterized by more extreme climatic conditions, boasting cold temperatures and short growing seasons. Resulting reduced primary production limits food availability and as ptarmigan are income breeders that use food resources acquired from nesting areas to supply energy and nutrients for egg production and incubation [@sandercock2005], lower carrying capacity in such areas is to be expected.

@@

Notably, the conclusive estimation of overall positive effects of rodent occupancy on recruitment in our model was only possible thanks to the integration and sharing of data across multiple areas. When @nilsen2023 fit the IDSM to data from only a single area, they were unable to obtain a reliable estimate for the rodent effect due to limited statistical power. Consequently, the space-for-time substitution that comes with extending the model across multiple area allows estimation of covariate effects that otherwise cannot be estimated, and opens up for future possibilities for studying effects of not just rodents, but also other environmental drivers on ptarmigan population dynamics. Doing so may also help with better understanding the mechanisms underlying the large portion of demographic rate variation that could only be attributed to random variation so far. This is the case especially for the relatively large residual variation ([@fig-varDecomp]) but also relevant for constant spatial and shared temporal variation. In previous work based on both marked [@Eriksen2023] and unmarked birds [@bowler2020; @Henden2020; @Novoa2016], spring conditions has come out as an important predictor of ptarmigan recruitment rates. In general, warmer and earlier springs seem to favour earlier breeding, larger clutch sizes [@Eriksen2023], and resulting higher recruitment rates measured in the late summer and early fall. @bowler2020 further reported that the strength of this relationship was not consistent in time and space, but was generally stronger in colder areas, similar to what we found for the effect of rodent occupancy here. In practice, measures representing spring conditions, such as the cover of ericaceous shrubs (a proxy for food availability) or spatially-explicit spring green up dates derived from remote-sensing data, thus constitute relevant candidate covariates for future work alongside temperature.

Another important determinant of vital rate variation is density dependence, in particular for exploited species like willow ptarmigan [@andrewartha1954; @sandercock_etal2011; @aanes2002; @willebrand2001]. Negative density dependence has been found in several gallinaceous birds such as northern bobwhites *Colinus virginianus* [@mcconnell2018], *Perdix perdix* [@bro2003], and wild turkeys *Meleagris gallopavo* [@mcghee2007]). For willow ptarmigan, evidence for density-dependent population regulation has been mixed. @myrberget1988, for example, observed no change in productivity despite a 50% decrease in abundance, while @pedersen2004 reported strong negative density-dependence over winter and posited that dispersal may be the vital rate that responded to changes in density most strongly. Similarly, @Henden2020 reported negative density dependence when using a Gompertz-model to examine how density and a range of environmental covariates affected willow ptarmigan population dynamics in the northernmost parts of Norway. While we did not explicitly model density dependence in this study, our results can provide some preliminary insights into potential density feedbacks from both a spatial (cross-population) and a temporal (within-population) angle. Comparing average population densities and growth rate across areas did not provide evidence for strong density dependence, but there was a tendency towards the highest population growth rates appearing in areas with relatively low density, and relatively low growth rates in high-density areas [@fig-paramCorrs A]. When considering density dependence across years within select areas, however, we found that higher density years were associated with higher recruitment the same year, but followed by lower apparent survival probabilities and, consequently, lower population growth rates (as determined by post-hoc Pearson correlation coefficients, supplementary file "DD_corrCoef.csv"). While this seems to support the notion of negative density-dependence, testing for this post-hoc gives results that are confounded with sampling correlation [@Freckleton2006]. Hence, formally modelling density-dependence, possibly using different forms and time-lags, could prove to be a promising extension of our modelling framework in the future.

@@

Producing a transparent and reproducible workflow for the analysis presented here was a central objective in this study. We have done this by setting up a well documented, function-based R workflow that allows (re-)running the complete analysis from downloading the publicly available data to visualizing the results produced by the IDSM [@fig-workflow] and that can be implemented using pipelines that employ "R targets" [@landau2021] and Nix/GNUparallel [@nix2004; @tange2024]. Modern applied ecology needs research to be published not just as scientific papers, but as reproducible and well documented workflows [@lewis2018]. This is particularly crucial for research that is (to be) closely tied to management and/or used to create biodiversity indicators that are to be reported nationally or internationally, or to be used by industrial partners [@powers2019]. This is both because of the enhanced transparency and credibility provided by openly available reproducible workflows and because of their cost-effectiveness, which allows for more sustainable use of funding in the mid- to long-term. Finally, open and reproducible workflows facilitate collaboration and inclusion of stakeholders in the research process, paving the path for the translational science that is required for society to tackle the the biodiversity crisis [@rubert2021]. It is our hope that this study can serve as an example of where to start.

@@

>>>>

---
title: "Large-scale spatiotemporal variation in vital rates and population dynamics of an alpine bird"
format: html
author: Chloé R. Nater, Francesco Frassinelli, James A. Martin, Erlend B. Nilsen
editor: visual
bibliography: rypeIDSM_MS2.bib
---

@@

Reproducibility and ease of repeating analyses was a key focus when developing the multi-area IDSM. To that end, we built a function-based workflow (@fig-workflow) that includes a variety of options for controlling modelling decisions such as the year range of data to consider, the level of spatial aggregation (i.e. reporting district vs. survey locality), whether to model time variation in survival and/or effects of rodent occupancy, whether to run MCMC chains sequentially or in parallel, etc. We then set up three pipelines for executing the workflow that differ in their degree of automation and user interface to meet different needs and resource constraints. The first is a step-by-step R script to be run manually that is suitable for exploration, development, and debugging. The second is a largely automated "R targets" pipeline [@landau2021], which allows executions through a single command and maximizes efficiency by keeping track of the "up-to-date" vs. "outdated" status of different steps in the workflow. The third is a pipeline that is run directly from command line, sets up a fully reproducible environment through Nix \[nix2004\], and parallelizes the MCMC outside of R using GNUparallel [@tange2024]. This option avoids a range of issues that can arise with R's internal parallelization (e.g. processes running even when the parent R session has been restarted, hard to debug, bad error handling, etc.) and is particularly well suited for running on servers and high-performance computing clusters. For more details on pipeline implementations and options, we refer the reader to the vignettes in the GitHub repository: <https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp>.

@@

```{r, echo = FALSE}
DDcorrs <- readRDS("Results/DD_corrCoef.rds")
DD_min_lam <- round(min(subset(DDcorrs, Parameter == "Lambda")$`50%`), digits = 3)
DD_max_lam <- round(max(subset(DDcorrs, Parameter == "Lambda")$`50%`), digits = 3)

VRcorrs <- readRDS("Results/VR_corrCoef.rds")
samCorr_mean <- round(subset(VRcorrs, area == "Overall")[,2:3], digits = 3)
```
Average population growth rates over the last four years (2018-2021) ranged from moderate declines (`r lambda.min$post.sum` in the "`r lambda.min$area.name`" area) to \> 50% increase (`r lambda.max$post.sum` in the "`r lambda.max$area.name`" area). In the majority of reporting areas (`r n.pos.lambda` out of 41), populations of willow ptarmigan have been increasing over the period 2018-2021 (@fig-lam). Some areas -- predominantly in central Norway -- also had declining populations, but many of those declines followed upon periods of increase between the start of data collection in 2007 and sometime between 2016 and 2018 (SF "TimeSeries_popDens1.pdf").

@@

The highest recent population growth rates were estimated for areas with relatively low recent population densities across latitudes but we did not find evidence for a strong association between population growth rates and population densities across areas in general ([@fig-paramCorrs] A). Within areas, however, we found substantial negative relationships between population density and population growth rates, with median correlation coefficients ranging from `r DD_max_lam` to `r DD_min_lam` (supplementary file "DD_corrCoef.csv").

@@

## Survival probabilities, recruitment rates, and generation times

@@

The MCMC chains for many of the area-specific average survival probabilities and recruitment rates, as well as for the global averages for both vital rates, were mixing rather poorly. Despite that, mixing was good and resulting posteriors well defined for the area- and year-specific estimates of survival and recruitment (SF "PostDens_tS_tR.pdf"). There was substantial variation in both vital rates across time (SFs "TimeSeries_pSurv.pdf" and "TimeSeries_rRep.pdf"). In a substantial number of areas, the years 2011, 2014, and 2018 not only supported high population densities (see above) but were also characterized by both high recruitment and low subsequent survival. The overall low density years 2012 and 2015, conversely, often featured lower recruitment and, in some cases, higher survival. Notably, there were also years with very little spatial synchrony, i.e. very different relative yearly survival probabilities and recruitment rates (e.g. 2010 and 2020 for survival and 2013, 2016, and 2017 for recruitment). This same pattern was also reflected in the within-sample correlations between population density in vital rates, which were predominantly positive for recruitment and negative for subsequent survival (supplementary file "DD_corrCoef.csv"). 
Sampling correlation between annual recruitment rates and survival probabilities was moderate when no time lag was considered ($R_t$ vs. $S_t$, average coefficient = `r samCorr_mean[1,2]`) and very low when comparing survival to subsequent recruitment ($R_t$ vs. $S_t$, average coefficient = `r samCorr_mean[2,2]`). Correlation coefficients varied substantially across areas though, featuring both positive and negative values with no clear spatial pattern (SF "SurvRepCorr_Latitude.pdf", supplementary file "VR_corrCoef.csv").

```{r, echo = FALSE}
GT <- readRDS("Results/PosteriorSummaries_GenTime_byArea.rds")
GT_R0 <- subset(GT, GT_measure == "Per generation growth rate")
GT_elas <- subset(GT, GT_measure == "Inverse of fecundity elasticity")
medianCorr <- round(cor(GT_R0$Median, GT_elas$Median), digits = 3)

minIdx <- which(GT$Median == min(GT$Median))
maxIdx <- which(GT$Median == max(GT$Median))

GT_min <- round(GT[minIdx, c("Median", "lCI", "uCI")], digits = 3)
GT_max <- round(GT[maxIdx, c("Median", "lCI", "uCI")], digits = 3)

GT_min_print <- paste0(GT_min[1], ", [", GT_min[2], ", ", GT_min[3], "]")
GT_max_print <- paste0(GT_max[1], ", [", GT_max[2], ", ", GT_max[3], "]")
```

Based on estimates of population growth rates and vital rates, we also calculated generation time as both per-generation population growth rate ($R0$) and inverse of fecundity elasticity ($1/elas_F$). The two approaches yielded very similar estimates (median correlation coefficient = `r medianCorr`) ranging from `r GT_min_print` to `r GT_max_print` years across areas (SF "GenerationTime_Latitude.pdf"). Spatial patterns in generation time were consistent with those for survival and recruitment, with the highest values occurring in the North and in the mountaineous regions in central Norway (SFs "GenerationTime_R0_Map.pdf" and "GenerationTime_elasF_Map.pdf").

@@

The model predicted a positive global effect of rodent occupancy on recruitment rate (average slope on the log scale = `r betaR.hyper.mean$post.sum`). Nonetheless, spatial variation in the rodent effect was non-negligible (random effect SD = `r betaR.sd.spatial$post.sum`). This resulted in negative (median) effects in `r n.neg.betaR` areas, positive (median) effects in `r n.pos.betaR` areas, and a range of effect sizes from `r betaR.min$post.sum` (area "`r betaR.min$area.name`") to `r betaR.max$post.sum` (area "`r betaR.max$area.name`", @fig-paramCorrs C, SF "Rep_betaR.R.pdf"). The largest positive rodent effects were estimated for areas in the very North of Norway, as well as in the mountaineous regions in the central and southwestern parts of the country (SF "betaR_Map.pdf"). Effects with negative posterior medians were located mostly at intermediate latitudes, but we note that all of these had posterior distributions featuring substantial overlap with 0 (@fig-paramCorrs C).

@@

Detection decay parameters, which determine detection probability in distance sampling surveys, varied across areas from between `r Det.min$post.sum` in "`r Det.min$area.name`" to `r Det.max$post.sum` in "`r Det.max$area.name`, resulting in detection probabilities over the transect sites ranging from `r pDet.min$post.sum` to `r pDet.max$post.sum`, respectively (truncation distance = 200 m) . The global average detection decay was `r Det.hyper.mean$post.sum` (detection probability = `r pDet.hyper.mean$post.sum`), and in general, higher values were more common in the Southern half of the country than the Northern half (SF"Avg_detect_Map.pdf"). Variation in detection over time was modest on average but the degree of temporal changes varied by area, with some areas having nearly constant detection while others showed variation by factors larger than 1.5 (SF "TimeSeries_pDetect.pdf"). The estimated average among-year variation in detection decay (random effect log SD = `r Det.sd.temporal$post.sum`) was lower than spatial (`r Det.sd.spatial$post.sum`) and residual (`r Det.sd.residual$post.sum`) variation.

@@

The relative importance of different components for explaining parameter variation differed among recruitment rate, survival probability, and detection decay (@fig-varDecomp). The largest portion of variation in recruitment was attributed to residual variation (`r Rec_R` %), followed by spatial (`r Rec_A` %) and temporal (`r Rec_Y` %) variation. Rodent occupancy, which contains both a spatial and a temporal dimension, explained `r Rec_Cov` % of the total variation. For survival, there was large uncertainty in the estimated proportions of variance explained by different components. The model predicted similar potential contributions from temporal (`r Surv_Y` %) and residual (`r Surv_R` %) variation and suggested that spatial variation was only responsible for `r Surv_A` % of the total variance. The majority of variance in detection decay was attributed evenly to spatial and residual variation at `r Det_A` % and `r Det_R` %, respectively. Temporal factors only accounted for `r Det_Y` %, of detection variation.

@@

Ptarmigan densities across the 41 reporting districts included in our analyses varied from around 2 birds/km^2^ to 55 birds/km^2^, with the lowest densities occurring far north in the country, as well as on the west coast and in the mountains in central Norway [@fig-densA]. The same spatial pattern was also evident at the level of the demographic rates: consistent with basic life history theory [@stearns1992], average recruitment rates were inversely related to average survival probabilities [@fig-paramCorrs B], and the slower life histories (higher survival, lower recruitment, and longer generation times) were more common in the northern and mountaineous parts of the country. This aligns with previous studies reporting relatively slower bird life histories in alpine / high altitude areas [e.g. @sandercock2005; @bears2009; @Wilson_Martin_2011; @alice2016]. In Norway, the northern and mountainous areas are characterized by more extreme climatic conditions, boasting cold temperatures and short growing seasons. Resulting reduced primary production limits food availability and as ptarmigan are income breeders that use food resources acquired from nesting areas to supply energy and nutrients for egg production and incubation [@sandercock2005], lower carrying capacity in such areas is to be expected.

@@

Notably, the conclusive estimation of overall positive effects of rodent occupancy on recruitment in our model was only possible thanks to the integration and sharing of data across multiple areas. When @nilsen2023 fit the IDSM to data from only a single area, they were unable to obtain a reliable estimate for the rodent effect due to limited statistical power. Consequently, the space-for-time substitution that comes with extending the model across multiple area allows estimation of covariate effects that otherwise cannot be estimated, and opens up for future possibilities for studying effects of not just rodents, but also other environmental drivers on ptarmigan population dynamics. Doing so may also help with better understanding the mechanisms underlying the large portion of demographic rate variation that could only be attributed to random variation so far. This is the case especially for the relatively large residual variation ([@fig-varDecomp]) but also relevant for constant spatial and shared temporal variation. In previous work based on both marked [@Eriksen2023] and unmarked birds [@bowler2020; @Henden2020; @Novoa2016], spring conditions have come out as an important predictor of ptarmigan recruitment rates. In general, warmer and earlier springs seem to favour earlier breeding, larger clutch sizes [@Eriksen2023], and resulting higher recruitment rates measured in the late summer and early fall. @bowler2020 further reported that the strength of this relationship was not consistent in time and space, but was generally stronger in colder areas, similar to what we found for the effect of rodent occupancy here. In practice, measures representing spring conditions, such as the cover of ericaceous shrubs (a proxy for food availability) or spatially-explicit spring green up dates derived from remote-sensing data, thus constitute relevant candidate covariates for future work alongside temperature.

Another important determinant of vital rate variation is density dependence, in particular for exploited species like willow ptarmigan [@andrewartha1954; @sandercock_etal2011; @aanes2002; @willebrand2001]. Negative density dependence has been found in several gallinaceous birds such as northern bobwhites *Colinus virginianus* [@mcconnell2018], *Perdix perdix* [@bro2003], and wild turkeys *Meleagris gallopavo* [@mcghee2007]). For willow ptarmigan, evidence for density-dependent population regulation has been mixed. @myrberget1988, for example, observed no change in productivity despite a 50% decrease in abundance, while @pedersen2004 reported strong negative density-dependence over winter and posited that dispersal may be the vital rate that responded to changes in density most strongly. Similarly, @Henden2020 reported negative density dependence when using a Gompertz-model to examine how density and a range of environmental covariates affected willow ptarmigan population dynamics in the northernmost parts of Norway. While we did not explicitly model density dependence in this study, our results can provide some preliminary insights into potential density feedbacks from both a spatial (cross-population) and a temporal (within-population) angle. Comparing average population densities and growth rate across areas did not provide evidence for strong density dependence, but there was a tendency towards the highest population growth rates appearing in areas with relatively low density, and relatively low growth rates in high-density areas [@fig-paramCorrs A]. When considering density dependence across years within select areas, however, we found that higher density years were associated with higher recruitment the same year, but followed by lower apparent survival probabilities and, consequently, lower population growth rates (as determined by post-hoc Pearson correlation coefficients, supplementary file "DD_corrCoef.csv"). While this seems to support the notion of negative density-dependence, testing for this post-hoc gives results that are confounded with sampling correlation [@Freckleton2006]. Our tests showed a moderate degree of sampling correlation between survival and recruitment on average (up to `r min(samCorr_mean)`), but there was substantial variation in the degree of correlation across areas (SF "SurvRepCorr_Latitude.pdf"). Hence, formally modelling density-dependence, possibly using different forms and time-lags, could prove to be a promising extension of our modelling framework in the future.

@@

Producing a transparent and reproducible workflow for the analysis presented here was a central objective in this study. We have done this by setting up a well documented, function-based R workflow that allows (re-)running the complete analysis from downloading the publicly available data to visualizing the results produced by the IDSM ([@fig-workflow]) and that can be implemented using pipelines that employ "R targets" [@landau2021] and Nix/GNUparallel [@nix2004; @tange2024]. Modern applied ecology needs research to be published not just as scientific papers, but as reproducible and well documented workflows [@lewis2018]. This is particularly crucial for research that is (to be) closely tied to management and/or used to create biodiversity indicators that are to be reported nationally or internationally, or to be used by industrial partners [@powers2019]. This is both because of the enhanced transparency and credibility provided by openly available reproducible workflows and because of their cost-effectiveness, which allows for more sustainable use of funding in the mid- to long-term. Finally, open and reproducible workflows facilitate collaboration and inclusion of stakeholders in the research process, paving the path for the translational science that is required for society to tackle the the biodiversity crisis [@rubert2021]. It is our hope that this study can serve as an example of where to start.

@@

**Francesco Frassinelli:** Software, Writing - Review and editing.
